---
title: "Leukocyte Composition Analysis of Combined DNHS+GTP 450K Cell Estimates - non-parametric, sex-stratified analyses for PTSDStatus"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---

```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(GGally)
# library(data.table)
library(psych)
library(rcompanion)
library(faraway)      # glm support
library(car)          # regression functions (continuous)
library(sandwich) ## heteroskedasticity robust standard errors.  
library(lmtest)   ## lm test
library(nortest) ## normality test
## further list of useful libraries in RegDiagNotes.Rmd

## further list of useful libraries in RegDiagNotes.Rmd
library(MASS)
library(coin)
library(Hmisc)
# library(lawstat) ## for levene.test to do pval adj

```

This notebook loads fully processed combined DNHS+GTP 450K phenotype data that includes all cell estimates based on *EpiDISH*'s implementation of robust partial correlation (RPC) generated in the **Combined\_450K\_BatchAdjsepwComBat\_Estimates\_2.Rmd** notebook in data processing repo.  These cell estimates were calculated after combining the datasets. 

```{r loadData}

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

source("../ModelFxns.R")
source("../PlotFxns.R")

rm(CombPheno2wcue2_tuk)

```

Use untransformed data for non-parametric tests. 

```{r out_dir}

proj_label <- "Combined_LeukoComp_450K_nonParametric_SexStrat_PTSDStatus_MO"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

# if (!file.exists(out_dir)) {
# dir.create(out_dir,showWarnings = FALSE,recursive = FALSE)
# }

```


```{r sub-bysex}

cells <- CombPheno2wcue2 %>% dplyr::select(c(starts_with("RPC_"), starts_with("CP_"))) %>% colnames()

# rpc <- CombPheno2wcue2 %>% dplyr::select(starts_with("RPC_")) %>% colnames()

CombPheno2wcue2 <- CombPheno2wcue2 %>% filter(!is.na(PTSDlife)) %>% mutate_at(vars(cells), funs(.*100)) %>% dplyr::select(-"Smoking") %>% dplyr::rename(ancPC1 = Comp.1, ancPC2 = Comp.2, Smoking = Smoking_pm)

CombPheno2wcue2_m <- CombPheno2wcue2 %>% filter(Sex == "Male")
CombPheno2wcue2_f <- CombPheno2wcue2 %>% filter(Sex == "Female")

```

```{r fxn-nonparatests}

nonParaTests <- function(data, testvar, cellset){
  
  testset <- lapply(cellset, function(x){
    
    lvlfctr <- levels(data[,testvar])
    form <- data[,x] ~ data[,testvar]
    
    if (length(lvlfctr) == 2) {
      cat("\n \n Checking normality, variance, and differences between groups for ", x, ". \n")
      print(shapiro.test(data[,x]))
      print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[1]]]))
      print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[2]]]))
      print(leveneTest(form, data = data))

    wt <- wilcox.test(form, data = data, paired = FALSE)
    print(wt)
    wt2 <- coin::wilcox_test(form, data = data, distribution = "exact")
    print(wt2)

list <- list(wt = wt, wt2 = wt2)
return(list)
    
    }
    
    if (length(lvlfctr) > 2) {
      cat("\n \n Checking normality, variance, and differences between groups for ", x, ". \n")
    print(shapiro.test(data[,x]))
    print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[1]]]))
    print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[2]]]))
    print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[3]]]))
    print(leveneTest(form, data = data))

    wt <- kruskal.test(form, data = data)
    print(wt)
    
    list <- list(wt = wt)
    }
    
  return(list)
  })
  
pvalwt <- vector(length = length(testset))

  for (i in 1:length(testset)) {
  pvalwt[i] <- testset[[i]]$wt$p.value
  }
method <- c("holm", "none")

pvaladj <- lapply(method, function(x){
  print(x)
  pvaladj_kw <- p.adjust(unlist(pvalwt), method = x)
print(pvaladj_kw) 

return(pvaladj_kw)
})

list2 <- list(testset = testset, pvaladj = pvaladj)

  return(list2)
}

```




## PTSD Status 

```{r CombPheno2wcue3}

CombPheno2wcue3 <- CombPheno2wcue2 %>% filter(!is.na(PTSDStatus))

CombPheno2wcue3_m <- CombPheno2wcue3 %>% filter(Sex == "Male")

CombPheno2wcue3_f <- CombPheno2wcue3 %>% filter(Sex == "Female")

table(CombPheno2wcue3$PTSDStatus, CombPheno2wcue3$Sex)

```


### RPC

```{r rpc-male-ptsdstatus}

rpc_male_PTSDStatus <- nonParaTests(data = CombPheno2wcue3_m, testvar = "PTSDStatus", cellset = "RPC_Mono")

```

Significant finding only in RPC monocytes in males.  

```{r rpc-female-ptsdstatus}

rpc_female_PTSDStatus <- nonParaTests(data = CombPheno2wcue3_f, testvar = "PTSDStatus", cellset = "RPC_Mono")

```

Nothing in females. 

### CP Monocytes

```{r cp-male-ptsdstatus}

cp_male_PTSDStatus <- nonParaTests(data = CombPheno2wcue3_m, testvar = "PTSDStatus", cellset = "CP_Mono")

```

Finding in CP monocytes in males.       

```{r cp-female-ptsdstatus}

cp_female_PTSDStatus <- nonParaTests(data = CombPheno2wcue3_f, testvar = "PTSDStatus", cellset = "CP_Mono")

```


### Post-hoc Dunn tests for both monocyte estimates in males

```{r dunntest}

library(FSA)

mono <- c("RPC_Mono", "CP_Mono")

## both sexes

dunn_mono_m <- lapply(mono, function(xx){
    cat("\n \n post-hoc Dunn test in males for ", xx, " by PTSDStatus.. \n")
    mo_kw <- dunnTest(CombPheno2wcue3_m[, xx] ~ CombPheno2wcue3_m$PTSDStatus, method = "holm")
    print(mo_kw)
    
  return(mo_kw)
})


```

Basically drives the same point.  
Difference in males is robust and seems to drive difference seen in both sexes.  

Significant comparisons were between control and both remitted and current PTSD, suggesting that lifetime PTSD status is driving difference independent of current state of PT symptoms.  

## Effect size of male monocyte findings

regarding calculating effect size for non-parametric tests.
  - can calculate for any test that returns Z-score.
  - turn into percentage variance measure of r by ... r = Z/sqrt(N), N = total num of observations.  
  - sign doesn't give helpful info so use abs value... 
  - r = 0.09 , example of small effect size.
  - estimate percentage of variance explained by squaring r.
  - r-family of effect size, which is percentage variance effect size.
  - different from d-family of effect size.

[reference for effect size here](http://documents.routledge-interactive.s3.amazonaws.com/9781138024571/supp/Classic_Non-Parametric_Statistics.JLH.pdf)

```{r effect-size}

nonParamEffectR <- function(Z, n){
  r = Z/sqrt(n)
  # print(r)
  return(r)
}

```

```{r effectsize-dunntest-PTSDStatus}

tab_m <- table(CombPheno2wcue3_m$PTSDStatus)
tab_m

## dunnTest for PTSDStatus

cat("\n effect size r-family, percentage variance, for RPC monocytes \n")
nonParamEffectR(dunn_mono_m[[1]]$res$Z[[1]], tab_m[[1]] + tab_m[[3]]) ## current vs controls. males. rpc.
nonParamEffectR(dunn_mono_m[[1]]$res$Z[[3]], tab_m[[1]] + tab_m[[2]]) ## remitted vs controls. males. rpc. inverse sign.
nonParamEffectR(dunn_mono_m[[1]]$res$Z[[2]], tab_m[[3]] + tab_m[[2]]) ## current vs. remitted PTSD.

cat("\n effect size r-family, percentage variance, for CP monocytes \n")
nonParamEffectR(dunn_mono_m[[2]]$res$Z[[1]], tab_m[[1]] + tab_m[[3]]) ## current vs controls. males. 
nonParamEffectR(dunn_mono_m[[2]]$res$Z[[3]], tab_m[[1]] + tab_m[[2]]) ## remitted vs controls. males. inverse sign. 
nonParamEffectR(dunn_mono_m[[2]]$res$Z[[2]], tab_m[[3]] + tab_m[[2]]) ## current vs. remitted PTSD.

```

## End

```{r save-end}

sessionInfo()

```
