---
title: "Leukocyte Composition Analysis of Combined DNHS+GTP 450K Cell Estimates - non-Parametric - calculate IQR and median"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---

```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(GGally)
# library(data.table)
library(psych)
library(rcompanion)
library(faraway)      # glm support
library(car)          # regression functions (continuous)
library(sandwich) ## heteroskedasticity robust standard errors.  
library(lmtest)   ## lm test
library(nortest) ## normality test
## further list of useful libraries in RegDiagNotes.Rmd

## further list of useful libraries in RegDiagNotes.Rmd
library(MASS)
library(coin)
library(Hmisc)
# library(lawstat) ## for levene.test to do pval adj

```

This notebook loads fully processed combined DNHS+GTP 450K phenotype data that includes all cell estimates based on *EpiDISH*'s implementation of robust partial correlation (RPC) generated in the **Combined\_450K\_BatchAdjsepwComBat\_Estimates\_2.Rmd** notebook in data processing repo.  These cell estimates were calculated after combining the datasets. 

```{r loadData}

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

rm(CombPheno2wcue2_tuk)

```

Use untransformed data for non-parametric tests. 

```{r out_dir}

proj_label <- "Combined_LeukoComp_450K_nonParametric"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

```


```{r sub-bysex}

cells <- CombPheno2wcue2 %>% dplyr::select(c(starts_with("RPC_"), starts_with("CP_"))) %>% colnames()

rpc <- CombPheno2wcue2 %>% dplyr::select(starts_with("RPC_")) %>% colnames()
cp <- CombPheno2wcue2 %>% dplyr::select(starts_with("CP_")) %>% colnames()

CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate_at(vars(cells), funs(.*100)) %>% dplyr::select(-"Smoking") %>% dplyr::rename(ancPC1 = Comp.1, ancPC2 = Comp.2, Smoking = Smoking_pm)

dim(CombPheno2wcue2)
CombPheno2wcue3 <- CombPheno2wcue2 %>% filter(!is.na(PTSDlife))

```

```{r}

# rpc
## summary stats for RPC cell estimates by sex.  
## standard median (not interpolated)
## trimmed means (drop top/bottom 0.1 fraction)
## type 3 skew/kurtosis
### Type 3 is b1 = [(n-1)/n]^{3/2} m_3/m_2^{3/2} and b2 = [(n-1)/n]^{3/2} m_4/m_2^2).

rpc_bysex_descStats <- sapply(rpc, function(x){
bysex <- psych::describeBy(CombPheno2wcue2[,x], IQR = TRUE, group = CombPheno2wcue2$Sex)
# print(bysex)
# male <- psych::describe(CombPheno2wcue2_m[,x], IQR = TRUE)
# print(male)
# list = list(fem = fem, male = male)
# return(list)
})
names(rpc_bysex_descStats) <- rpc

rpc_bysex_descStats

```

```{r}

sapply(rpc, function(x){summary(CombPheno2wcue2_m[,x])})
sapply(rpc, function(x){summary(CombPheno2wcue2_f[,x])})

```

```{r}

sapply(cells[7:12], function(x){summary(CombPheno2wcue2_m[,x])})
sapply(cells[7:12], function(x){summary(CombPheno2wcue2_f[,x])})

```

```{r}

# cp
## summary stats for RPC cell estimates by sex.  
## standard median (not interpolated)
## trimmed means (drop top/bottom 0.1 fraction)
## type 3 skew/kurtosis
### Type 3 is b1 = [(n-1)/n]^{3/2} m_3/m_2^{3/2} and b2 = [(n-1)/n]^{3/2} m_4/m_2^2).

cp_bysex_descStats <- lapply(cells[7:12], function(x){
psych::describeBy(CombPheno2wcue2[,x], IQR = TRUE, group = CombPheno2wcue2$Sex)
})
names(cp_bysex_descStats) <- cells[7:12]

cp_bysex_descStats

```