---
title: "DNHS and GTP comparison - Leuko Comp"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: default
---

```{r knitrOpt}

rm(list = ls())
library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70)
render_listings()

```


```{r load, include = FALSE, message = FALSE}

library(tidyverse) # data management/manipulation

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

```


```{r nonParaTests-fxn, include = FALSE, message = FALSE}

library(psych)
library(rcompanion)
library(faraway)      # glm support
library(car)          # regression functions (continuous)
library(sandwich) ## heteroskedasticity robust standard errors.  
library(lmtest)   ## lm test
library(nortest) ## normality test
library(MASS)
library(coin)
library(Hmisc)

nonParaTests <- function(data, testvar, cellset){
  
  testset <- lapply(cellset, function(x){
    
    lvlfctr <- levels(data[,testvar])
    form <- data[,x] ~ data[,testvar]
    
    if (length(lvlfctr) == 2) {
      cat("\n \n Checking normality, variance, and differences between groups for ", x, ". \n")
      print(shapiro.test(data[,x]))
      print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[1]]]))
      print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[2]]]))
      print(leveneTest(form, data = data))
    wt <- wilcox.test(form, data = data, paired = FALSE)
    print(wt)
    wt2 <- coin::wilcox_test(form, data = data, distribution = "exact")
    print(wt2)
list <- list(wt = wt, wt2 = wt2)
return(list)
    
    }
    
    if (length(lvlfctr) > 2) {
      cat("\n \n Checking normality, variance, and differences between groups for ", x, ". \n")
    print(shapiro.test(data[,x]))
    print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[1]]]))
    print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[2]]]))
    print(shapiro.test(data[,x][data[,testvar] == lvlfctr[[3]]]))
    print(leveneTest(form, data = data))
    wt <- kruskal.test(form, data = data)
    print(wt)
    
    list <- list(wt = wt)
    }
    
  return(list)
  })
  
pvalwt <- vector(length = length(testset))
  for (i in 1:length(testset)) {
  pvalwt[i] <- testset[[i]]$wt$p.value
  }
method <- c("holm", "none")
pvaladj <- lapply(method, function(x){
  print(x)
  pvaladj_kw <- p.adjust(unlist(pvalwt), method = x)
print(pvaladj_kw) 
return(pvaladj_kw)
})
list2 <- list(testset = testset, pvaladj = pvaladj)
  return(list2)
}

```

To be consistent with previous analyses conducted only on those with lifetime PTSD data, we redid non-parametric analyses for study on subset of 483 (instead of full set of 565, including those in GTP without lifetime PTSD data).  

```{r sub-study483}

CombPheno2wcue2 <- CombPheno2wcue2 %>% filter(!is.na(PTSDlife))
female <- subset(CombPheno2wcue2, Sex == "Female")
male <- subset(CombPheno2wcue2, Sex == "Male")

rpc <- CombPheno2wcue2 %>% dplyr::select(starts_with("RPC_")) %>% colnames()

```


```{r run-nonpara-rpc}

nonPT_Study <- nonParaTests(data = CombPheno2wcue2, testvar = "Study", cellset = rpc)

nonPTfem_Study <- nonParaTests(data = female, testvar = "Study", cellset = rpc)

nonPTmale_Study <- nonParaTests(data = male, testvar = "Study", cellset = rpc)


```

```{r run-nonpara-cp}

cp <- CombPheno2wcue2 %>% dplyr::select(starts_with("CP_")) %>% colnames()

nonPT_Study <- nonParaTests(data = CombPheno2wcue2, testvar = "Study", cellset = cp)

nonPTfem_Study <- nonParaTests(data = female, testvar = "Study", cellset = cp)

nonPTmale_Study <- nonParaTests(data = male, testvar = "Study", cellset = cp)


```


```{r padj-levene}

rpc_lev <- c(0.1776, 0.03248, 0.8327, 0.9412, 0.2698, 0.4579)
p.adjust(rpc_lev, method = "holm")

rpc_lev_male <- c(0.1048, 0.02287, 0.5239, 0.9381, 0.2412, 0.1528)
p.adjust(rpc_lev_male, method = "holm")



cp_lev <- c(0.1214, 0.03905, 0.2981, 0.9977, 0.3665, 0.05999)
p.adjust(cp_lev, method = "holm")

cp_lev_male <- c(0.09124, 0.02312, 0.1167, 0.874, 0.2262, 0.0252)
p.adjust(cp_lev_male, method = "holm")


```


```{r kstest-nomp-lev}

ks.test(subset(CombPheno2wcue2$RPC_NK, CombPheno2wcue2$Study == "GTP"), subset(CombPheno2wcue2$RPC_NK, CombPheno2wcue2$Study == "DNHS"))

ks.test(subset(male$RPC_NK, male$Study == "GTP"), subset(male$RPC_NK, male$Study == "DNHS"))

ks.test(subset(CombPheno2wcue2$CP_NK, CombPheno2wcue2$Study == "GTP"), subset(CombPheno2wcue2$CP_NK, CombPheno2wcue2$Study == "DNHS"))

ks.test(subset(male$CP_NK, male$Study == "GTP"), subset(male$CP_NK, male$Study == "DNHS"))


ks.test(subset(male$CP_Mono, male$Study == "GTP"), subset(male$CP_Mono, male$Study == "DNHS"))

```

KS test not significant for any not equal in variance at nominal p. 

other

```{r padj-bysex-lev}

rpc_lev_sex <- c(0.29, 8.1e-05, 0.42, 0.0013, 0.076, 0.077)
p.adjust(rpc_lev_sex, method = "holm")

cp_lev_sex <- c(0.38, 1e-04, 0.66, 0.00016, 0.1, 0.054)
p.adjust(cp_lev_sex, method = "holm")

```

## Effect size of findings

regarding calculating effect size for non-parametric tests.
  - can calculate for any test that returns Z-score.
  - turn into percentage variance measure of r by ... r = Z/sqrt(N), N = total num of observations.  
  - sign doesn't give helpful info so use abs value... 
  - r = 0.09 , example of small effect size.
  - estimate percentage of variance explained by squaring r.
  - r-family of effect size, which is percentage variance effect size.
  - different from d-family of effect size.

[reference for effect size here](http://documents.routledge-interactive.s3.amazonaws.com/9781138024571/supp/Classic_Non-Parametric_Statistics.JLH.pdf)


```{r effect-size}

nonParamEffectR <- function(Z, n){
  r = Z/sqrt(n)
  # print(r)
  return(r)
}

## RPC
# Z = 2.9
# N = 153
# rpc_r_mw <- nonParamEffectR(2.9, 153)
# rpc_r2 <- rpc_r_mw^2
# 
# nonParamEffectR(0.58, 330)
# 
# ## cp
# nonParamEffectR(2.5, 153)
# nonParamEffectR(0.15, 330)

```

```{r eff-size-calc}
## combined. 
nonParamEffectR(-2.9272, 483)
nonParamEffectR(-2.856, 483)

## male
nonParamEffectR(-3.24, 153)

## GTP
# nonParamEffectR(-1.4197, 89)

```

```{r study-tab-chisq}

chisq.test(table(CombPheno2wcue2$Race, CombPheno2wcue2$Study), simulate.p.value = TRUE)
chisq.test(table(CombPheno2wcue2$PTSDlife, CombPheno2wcue2$Study), simulate.p.value = TRUE)

table(CombPheno2wcue2$PTSDlife, CombPheno2wcue2$Study)
table(CombPheno2wcue2$Race, CombPheno2wcue2$Study)
table(CombPheno2wcue2$Sex, CombPheno2wcue2$PTSDlife, CombPheno2wcue2$Study)


table(CombPheno2wcue2$Sex, CombPheno2wcue2$PTSDlife)

```

## Plots 

```{r lib-plots, message = FALSE, include = FALSE}

library(GGally)
library(data.table)
library(ggpubr)
library(ggsci)
library(grid)
library(gtable)
library(gridExtra)
library(RColorBrewer)
library(scales)

```

```{r out_dir}

proj_label <- "added_Plots_rev1"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

if (!file.exists(out_dir)) {
dir.create(out_dir,showWarnings = FALSE,recursive = FALSE)
}

```

```{r CombPheno2wcue2-rmpheno}

cells <- CombPheno2wcue2 %>% dplyr::select(c(ends_with("_Mono"), ends_with("_Gran"), ends_with("_CD4T"), ends_with("_CD8T"), ends_with("_NK"), ends_with("_B", ignore.case = FALSE))) %>% dplyr::select(-starts_with("CBS_")) %>% colnames()

CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate_at(vars(cells), funs(.*100)) %>% filter(!is.na(PTSDlife)) 
CombPheno2wcue2
# CombPheno2wcue2 %>% filter(is.na(Age)) ## none missing age

```

- scale everything as percent.

```{r set-palette}

rpc_ct <- CombPheno2wcue2 %>% dplyr::select(starts_with("RPC_")) %>% colnames()
cp_ct <- CombPheno2wcue2 %>% dplyr::select(starts_with("CP_")) %>% colnames()

sexpalbrewed <- brewer.pal(8, "Dark2")
sexpal <- sexpalbrewed[c(2,3)]

ptsdpal <- c("#619CFF", "#DC0000E5") ## 
ptsdpal2 <- c("#619CFF", "#00BA38", "#E64B35E5") ## 

```

```{r nonp-cor}

cat("\n \n Kendall's correlation between RPC estimates and Age. \n")

rpc_cor_ke <- lapply(rpc_ct, function(x){
  
  cor <- cor.test(CombPheno2wcue2[,x], CombPheno2wcue2$Age, method = "kendall", continuity = TRUE)
  print(cor)
  
})

rpc_ke_pval <- sapply(1:6, function(x) rpc_cor_ke[[x]]$p.value)
p.adjust(rpc_ke_pval, method = "holm")


cat("\n \n Spearman's correlation between RPC estimates and Age. \n")

rpc_cor_sp <- lapply(rpc_ct, function(x){
  
  cor <- cor.test(CombPheno2wcue2[,x], CombPheno2wcue2$Age, method = "spearman", continuity = TRUE)
  print(cor)
  
})

rpc_sp_pval <- sapply(1:6, function(x) rpc_cor_sp[[x]]$p.value)
p.adjust(rpc_sp_pval, method = "holm")

cat("\n \n Kendall's correlation between CP estimates and Age. \n")

cp_cor_ke <- lapply(cp_ct, function(x){
  
  cor <- cor.test(CombPheno2wcue2[,x], CombPheno2wcue2$Age, method = "kendall", continuity = TRUE)
  print(cor)
  
})

cp_ke_pval <- sapply(1:6, function(x) cp_cor_ke[[x]]$p.value)
p.adjust(cp_ke_pval, method = "holm")

cat("\n \n Spearman's correlation between CP estimates and Age. \n")

cp_cor_sp <- lapply(cp_ct, function(x){
  
  cor <- cor.test(CombPheno2wcue2[,x], CombPheno2wcue2$Age, method = "spearman", continuity = TRUE)
  print(cor)
  
})

cp_sp_pval <- sapply(1:6, function(x) cp_cor_sp[[x]]$p.value)
p.adjust(cp_sp_pval, method = "holm")

```

```{r corcomp-age}

rpc_age <- lapply(rpc_ct, function(cell){
  lab <- gsub(cell, pattern = "_", replacement = " ")
  ggscatter(data = CombPheno2wcue2, x = "Age", y = cell, point = TRUE, add = "reg.line", conf.int = TRUE, size = 1, ylab = lab) +
  stat_cor(method = "spearman") 

})

cp_age <- lapply(cp_ct, function(cell){
  lab <- gsub(cell, pattern = "_", replacement = " ")
  ggscatter(data = CombPheno2wcue2, x = "Age", y = cell, point = TRUE, add = "reg.line", conf.int = TRUE, size = 1, ylab = lab) +
  stat_cor(method = "spearman") 

})


pdf(file = paste0(out_dir, "CorComp_Age_spearman_diffs.pdf"), width = 9, height = 5, title = paste0("CorComp_age_spearman"), bg = "transparent")

ggarrange(plotlist = rpc_age, ncol = 3, nrow = 2)
ggarrange(plotlist = cp_age, ncol = 3, nrow = 2)

dev.off()

```


```{r corcomp-age-bysex, eval = FALSE, include = FALSE}

rpc_ages <- lapply(rpc_ct, function(cell){
  lab <- gsub(cell, pattern = "_", replacement = " ")
  ggscatter(data = CombPheno2wcue2, x = "Age", y = cell, color = "Sex", palette = sexpal, point = TRUE, ylab = lab, add = "reg.line", conf.int = TRUE, size = 1) +
  stat_cor(aes(color = Sex), method = "spearman")

})

cp_ages <- lapply(cp_ct, function(cell){
  lab <- gsub(cell, pattern = "_", replacement = " ")
  ggscatter(data = CombPheno2wcue2, x = "Age", y = cell, color = "Sex", palette = sexpal, point = TRUE, ylab = lab, add = "reg.line", conf.int = TRUE, size = 1) +
  stat_cor(aes(color = Sex), method = "spearman")
  
})


pdf(file = paste0(out_dir, "CorComp_Age_spearman_diffs_bySex.pdf"), width = 9, height = 5, title = paste0("CorComp_age_spearman_diffs_bySex"), bg = "transparent")

ggarrange(plotlist = rpc_ages, ncol = 3, nrow = 2, legend = "top", common.legend = TRUE)
ggarrange(plotlist = cp_ages, ncol = 3, nrow = 2, legend = "top", common.legend = TRUE)

dev.off()

```
