---
title: "Phenotype Prep and Transformation of Combined DNHS+GTP 450K Cell Estimates"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---

```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(GGally)
library(data.table)
library(psych)
library(rcompanion)
library(faraway)      # glm support
library(car)          # regression functions (continuous)
library(sandwich) ## heteroskedasticity robust standard errors.  
library(lmtest)   ## lm test
library(nortest) ## normality test
## further list of useful libraries in RegDiagNotes.Rmd
library(moments)
library(e1071)                    

```

This notebook loads fully processed combined DNHS+GTP 450K phenotype data that includes all cell estimates calculated on the combined dataset, based on *EpiDISH* generated in the **Combined\_450K\_BatchAdjsepwComBat\_Estimates\_2.Rmd** notebook in data processing repo.  

```{r loadData}

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "Combined_450K_BatchAdjsepwComBat_Estimates.Rdata"))
load(file = "D://Data/Combined_450K_Pheno.Rdata")

```

```{r out_dir}

proj_label <- "Combined_PhenoPrep_Transform_0"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

# if (!file.exists(out_dir)) {
# dir.create(out_dir,showWarnings = TRUE,recursive = FALSE)
# }

```

Recode current PTSD Dx to try out GEO's for GTP. 

```{r switch-PTSDpm2-gtp}

CombPheno2wcue2 <- left_join(CombPheno2wcue, GTPpheno390wc4[, c("SampleID", "PTSD_Dx")], by = "SampleID")
CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate(PTSDpm2 = case_when(Study == "GTP" ~ PTSD_Dx, Study == "DNHS" ~ PTSDpm))

table(CombPheno2wcue2$PTSDpm, CombPheno2wcue2$PTSDpm2)

```

Much more informative. More relevant to DNHS, because DNHS is also self-report.  Something seems obscure about the CAPS current PTSD.

```{r addPTSDStatus2}

CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate(PTSDStatus2 = case_when(PTSDpm2 == "yes" ~ "Current", PTSDpm2 == "no" & PTSDlife == "yes" ~ "Remitted", PTSDlife == "no" ~ "No PTSD")) %>% mutate(PTSDStatus2 = factor(PTSDStatus2, levels = c("No PTSD", "Remitted", "Current")))

```

```{r cells-cknormality}

cells <- CombPheno2wcue2 %>% dplyr::select(c(starts_with("RPC_"), starts_with("CP_"), starts_with("CBS_"))) %>% colnames()

lapply(CombPheno2wcue2[,cells], shapiro.test)

```

only granulocytes pass normality test, and is trending

```{r ck-normal-log, eval = FALSE, include = FALSE}

lapply(CombPheno2wcue2[,cells], function(x) shapiro.test(log(x))) ## lot of NAs
lapply(CombPheno2wcue2[,cells], function(x) shapiro.test(log2(x)))

```

```{r cells-ckdensity, message = FALSE}

lapply(cells, function(x){
  ggplot(CombPheno2wcue2, aes_string(x = x, y = "..density..")) + geom_density() + ggtitle(x)
  
})


```

```{r cells-ckskewkurt-untransformed}
# To check skewness and kurtosis

cat("\n \n Check skewness before transform using 2 methods. \n ")

sapply(CombPheno2wcue2[,cells], moments::skewness)

sapply(CombPheno2wcue2[,cells], e1071::skewness)

cat("\n \n Check kurtosis before transform using 2 methods. \n ")

sapply(CombPheno2wcue2[,cells], moments::kurtosis) ## Pearson's measure of kurtosis

sapply(CombPheno2wcue2[,cells], e1071::kurtosis)

```

Cell Estimates are tukey-transformed for linear regression models.  

```{r transformTukey}

# cells_tuk <- c("B", "NK", "CD4T", "CD8T", "Mono", "Gran")

Combined_cells_tuk <- CombPheno2wcue2[,c("SampleID", cells)]

Combined_cells_tuk[,-1] <- sapply(Combined_cells_tuk[,-1], rcompanion::transformTukey)

CombPheno2wcue2_nocell <- CombPheno2wcue2 %>% dplyr::select(-c(one_of(cells)))

CombPheno2wcue2_tuk <- left_join(CombPheno2wcue2_nocell, Combined_cells_tuk, by = "SampleID")

# lapply(Combined_cells_tuk[,-1], shapiro.test)

```


```{r CenterVar}

CenterVar <- c("Comp.1", "Comp.2")

CombPheno2wcue2_tuk[,CenterVar] <- sapply(CombPheno2wcue2_tuk[,CenterVar], function(x){
  scale(x, center = TRUE, scale = FALSE)
})

sapply(CombPheno2wcue2_tuk[,CenterVar], summary)
sapply(CombPheno2wcue2_tuk[,CenterVar], densityPlot)

```

```{r transformed-densplots}

lapply(cells, function(x){
  ggplot(CombPheno2wcue2_tuk, aes_string(x = x, y = "..density..")) + geom_density() + ggtitle(x)
  
})

```

```{r cells-tuk-cksk2}

# To check skewness and kurtosis after transform

cat("\n \n Check skewness after Tukey transform using 2 methods. \n ")
sapply(CombPheno2wcue2_tuk[,cells], moments::skewness)

sapply(CombPheno2wcue2_tuk[,cells], e1071::skewness)

cat("\n \n Check kurtosis after Tukey transform using 2 methods. \n ")

sapply(CombPheno2wcue2_tuk[,cells], moments::kurtosis) ## Pearson's measure of kurtosis
sapply(CombPheno2wcue2_tuk[,cells], e1071::kurtosis)

## improved. w transformation
 
```

```{r ck-numna, eval = FALSE, include = FALSE}

lapply(CombPheno2wcue2_tuk, function(x){
  length(x[is.na(x)])
})

```

Improved, but not normality not fully addressed for all cell types.   
Save both untransformed and Tukey-transformed phenotype data.  

Note that Tukey-transformed phenotype data has ancestry PCs centered, but untransformed does not.  

Also note that we decided to just use RPC and CP cell estimates due to lack of confidence in CBS estimates, which occasionally deviates and has been also noted in Salas et al. (2018).

- Do plots (e.g., boxplots/violinplots(?) and density plots) on both untransformed and tukey-transformed data.  Can you see some patterns in Tukey-transformed that weren't visible in untransformed data?  What does this mean?
- ggduo? and strat by sex?
- Parametric and non-parametric tests.  

## Save

```{r sessionInfo}

save(CombPheno2wcue2_tuk, CombPheno2wcue2, file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

save.image(paste0(proj_label, ".RData"))

sessionInfo()

```
