---
title: "Leukocyte Composition Analysis of Combined DNHS+GTP 450K Cell Estimates - Table 1 - Descriptive Statistics"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---

```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(psych)
library(Gmisc)

```

```{r loadData}

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

rm(CombPheno2wcue2_tuk)

```

```{r out_dir}

proj_label <- "Combined_LeukoComp_450K_Tab1_rev1"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

# if (!file.exists(out_dir)) {
# dir.create(out_dir,showWarnings = FALSE,recursive = FALSE)
# }

```

Print out table 1 descriptive statistics

### Table 1 Descriptive Stats

```{r datatab1sel}

datatab1sel <- CombPheno2wcue2 %>% filter(!is.na(PTSDlife)) %>% dplyr::select(one_of("Study", "group", "Race", "Age", "Smoking_pm")) %>% dplyr::rename(Smoking = Smoking_pm)
varnames <- colnames(datatab1sel)[-1]

```

```{r getT1Stat-fxn}

library(Gmisc)
# A function that takes the variable name,
# applies it to the melanoma dataset
# and then runs the results by the status variable
getT1Stat <- function(varnames){
getDescriptionStatsBy(datatab1sel[, varnames],
datatab1sel$Study,
add_total_col = "last",
show_all_values = TRUE,
continuous_fn = describeMedian,
statistics = FALSE,
html = TRUE,
header_count = TRUE)
}

```

```{r tab-data}
# Save everything in a list
# This simplifies the row grouping
table_data <- list()
# Get the basic stats
table_data[["group"]] <- getT1Stat("group")
table_data[["Race"]] <- getT1Stat("Race")
table_data[["Median Age"]] <- getT1Stat("Age")
table_data[["Current Smoking"]] <- getT1Stat("Smoking")
# Now merge everything into a matrix
# and create the rgroup & n.rgroup variables
rgroup <- c()
n.rgroup <- c()
output_data <- NULL
for (varlabel in names(table_data)) {
output_data <- rbind(output_data,
table_data[[varlabel]])
rgroup <- c(rgroup,
varlabel)
n.rgroup <- c(n.rgroup,
nrow(table_data[[varlabel]]))
}
output_data

```

```{r htmlTab}

# options(htmlTable.cat = TRUE)
htmlTable(output_data, align = "lrr",
rgroup = rgroup, n.rgroup = n.rgroup,
rgroupCSSseparator = "",
rowlabel = "",
caption = "Key demographic characteristics of the DNHS and GTP",
tfoot = "This table describes the subset of participants included in primary analyses investigating sex-specific associations between DNAm-based cell estimates and lifetime PTSD.", ctable = TRUE)

```

html table printed out in this R notebook was used as Table 1 for this paper.  The table was copy and pasted onto word docx for additional formatting and addition of remaining footnote.  (only first sentence included here.)

## Table 1 alt

Key demographic characteristics by sex. 

```{r datatab1sel2}

datatab1sel2 <- CombPheno2wcue2 %>% filter(!is.na(PTSDlife)) %>% dplyr::select(one_of("Study", "Sex", "Race", "Age", "Smoking_pm","PTSDlife")) %>% dplyr::rename(PTSD = PTSDlife, Smoking = Smoking_pm)

varnames2 <- colnames(datatab1sel2)[-2]

```

```{r getT1Stat2-fxn}

library(Gmisc)
# A function that takes the variable name,
# applies it to the melanoma dataset
# and then runs the results by the status variable
getT1Stat2 <- function(varnames2){
getDescriptionStatsBy(datatab1sel2[, varnames2],
datatab1sel2$Sex,
add_total_col = "last",
show_all_values = TRUE,
continuous_fn = describeMedian,
statistics = FALSE,
html = TRUE,
header_count = TRUE)
}

```

```{r tab-data2}
# Save everything in a list
# This simplifies the row grouping
table_data2 <- list()
# Get the basic stats
table_data2[["Study"]] <- getT1Stat2("Study")
table_data2[["Race"]] <- getT1Stat2("Race")
table_data2[["Median Age"]] <- getT1Stat2("Age")
table_data2[["Current Smoking"]] <- getT1Stat2("Smoking")
table_data2[["Lifetime PTSD"]] <- getT1Stat2("PTSD")
# Now merge everything into a matrix
# and create the rgroup & n.rgroup variables
rgroup <- c()
n.rgroup <- c()
output_data2 <- NULL
for (varlabel in names(table_data2)) {
output_data2 <- rbind(output_data2,
table_data2[[varlabel]])
rgroup <- c(rgroup,
varlabel)
n.rgroup <- c(n.rgroup,
nrow(table_data2[[varlabel]]))
}
output_data2

```

```{r htmlTab2}

# options(htmlTable.cat = TRUE)
htmlTable(output_data2, align = "lrr",
rgroup = rgroup, n.rgroup = n.rgroup,
rgroupCSSseparator = "",
rowlabel = "",
caption = "Key demographic characteristics by Sex",
tfoot = "This table describes the subset of participants included in primary analyses investigating sex-specific associations between DNAm-based cell estimates and lifetime PTSD, by sex.", ctable = TRUE)

```

html table printed out in this R notebook was used as Table 1 for this paper.  The table was copy and pasted onto word docx for additional formatting and addition of remaining footnote.  (only first sentence included here.)


```{r save-end}

# save.image(paste0(proj_label, ".RData"))
sessionInfo()

```


