---
title: "Additional Figures and Plots - Combined DNHS+GTP 450K Cell Estimates - v3"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---


```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

# rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(GGally)
library(data.table)
library(ggpubr)
library(ggsci)
library(grid)
library(gtable)
library(gridExtra)
library(RColorBrewer)
library(scales)

```

Generates some additional plots and figures, not in original set generated in **Combined_LeukoComp_450K_FigPlots.Rmd**.  
Load saved image from original set.  

```{r loadData-old-image}

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

table(CombPheno2wcue2$PTSDlife, CombPheno2wcue2$Sex)

```

```{r out_dir}

proj_label <- "added_Figures_Plots_v3"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

if (!file.exists(out_dir)) {
dir.create(out_dir,showWarnings = FALSE,recursive = FALSE)
}

```


```{r CombPheno2wcue2-rmpheno}

cells <- CombPheno2wcue2 %>% dplyr::select(c(ends_with("_Mono"), ends_with("_Gran"), ends_with("_CD4T"), ends_with("_CD8T"), ends_with("_NK"), ends_with("_B", ignore.case = FALSE))) %>% dplyr::select(-starts_with("CBS_")) %>% colnames()

CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate_at(vars(cells), funs(.*100)) %>% filter(!is.na(PTSDlife)) 
CombPheno2wcue2

```

- final set. v3. 
- scale everything as percent.

```{r set-palette}

sexpalbrewed <- brewer.pal(8, "Dark2")
sexpal <- sexpalbrewed[c(2,3)]

ptsdpal <- c("#619CFF", "#DC0000E5") ## 
ptsdpal2 <- c("#619CFF", "#00BA38", "#E64B35E5") ## 

```


```{r pheno450k_allCE}

rpc <- CombPheno2wcue2 %>% dplyr::select(c(starts_with("RPC_"), matches("SampleID")))
colnames(rpc) <- gsub(colnames(rpc), pattern = "RPC_", replacement = "")

vars <- CombPheno2wcue2 %>% dplyr::select(-c(starts_with("RPC_"), starts_with("CBS_")))
colnames(vars) <- gsub(colnames(vars), pattern = "CP_", replacement = "")

vars.dtl <- melt(data.table(vars), 
                 # id.vars = "SampleID", 
                 measure.vars = c("B", "NK", "CD4T", "CD8T", "Gran", "Mono"), 
                 variable.name = "CellType", 
                 value.name = "CP")

rpc.dtl <- melt(data.table(rpc), 
                 id.vars = "SampleID", 
                 measure.vars = c("B", "NK", "CD4T", "CD8T", "Gran", "Mono"), 
                 variable.name = "CellType", 
                 value.name = "RPC")

setkey(rpc.dtl, SampleID, CellType)
setkey(vars.dtl, SampleID, CellType)

pheno450k_allCE <- merge(vars.dtl,rpc.dtl,by = c("SampleID","CellType"))
pheno450k_allCE

```

```{r suppfig1-pheno450k_allCE-plotcor-spearman}

pdf(file = paste0(out_dir, "SuppFig1Raw_CorComp_bySex_spearman.pdf"), width = 9, height = 5, title = paste0("SuppFig1_CorComp_spearman"), bg = "transparent")

ggscatter(data = pheno450k_allCE, x = "CP", y = "RPC", color = "Sex", palette = sexpal, point = TRUE, facet.by = "CellType", add = "reg.line", conf.int = TRUE, size = 1) +
  stat_cor(aes(color = Sex), method = "spearman") +
  geom_abline(slope = 1, intercept = 0) +
  theme(strip.text = element_text(size = 14, face = "bold"),    text = element_text(size = 14, face = "bold"),
legend.title = element_text(face = "bold"), axis.title = element_text(face = "bold"), axis.text = element_text(face = "plain", size = 12), legend.position = "right", axis.title.x = element_text(margin = margin(0.5, 0, 0, 0, unit = "cm")), axis.title.y = element_text(margin = margin(0, 0.5, 0, 0, unit = "cm")))
# , plot.tag = element_text(face = "bold")
dev.off()
        
```

Spearman makes the most sense. 
Spearman rank correlation: Spearman rank correlation is a non-parametric test that is used to measure the degree of association between two variables.  The Spearman rank correlation test does not carry any assumptions about the distribution of the data and is the appropriate correlation analysis when the variables are measured on a scale that is at least ordinal.



```{r fig1-suppfig2-densityPlots-fig1-ggpubr}

pdf(file = paste0(out_dir, "Fig1Raw_RPC_DensityPlot_byCellType.pdf"), width = 9, height = 5, title = paste0("Fig1Raw_RPC_Density_byCellType"), bg = "transparent")

p1 <- ggdensity(pheno450k_allCE, x = "RPC", fill = "Sex", color = "Sex", palette = sexpal, facet.by = "CellType", alpha = 0.5) + theme(strip.text = element_text(face = "bold", size = 12), legend.text = element_text(size = 11, face = "plain"), legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"), axis.text = element_text(size = 12, face = "plain"), legend.position = "bottom", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))

print(p1)

dev.off()

pdf(file = paste0(out_dir,"SuppFig2Raw_CP_DensityPlot_byCellType.pdf"), width = 9, height = 5, title = paste0("SuppFig2Raw_CP_DensityPlot_byCellType"), bg = "transparent")

p2d <- ggdensity(pheno450k_allCE, x = "CP", fill = "Sex", color = "Sex", palette = sexpal, facet.by = "CellType", alpha = 0.5) + theme(strip.text = element_text(face = "bold", size = 12), legend.text = element_text(size = 11, face = "plain"), legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"), axis.text = element_text(size = 12, face = "plain"), legend.position = "bottom", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))

print(p2d)

dev.off()

# removed options. add = mean, rug = TRUE

```



```{r fig2violinplot-RPC-ggpubr-fig2}

## add p-values comparing groups, via non-parametric Mann-Whitney U test. 

## non-parametric: Mann-Whitney U test
compare_means(formula = c(RPC_Gran, RPC_Mono, RPC_B, RPC_CD4T, RPC_CD8T, RPC_NK) ~ PTSDlife, data = CombPheno2wcue2, method = "wilcox.test", paired = FALSE, group.by = "Sex", ref.group = "no", p.adjust.method = "holm")

 pdf(file = paste0(out_dir, "Fig2Raw_RPCest_violinplots.pdf"),
    bg = "transparent",
    width = 9,
    height = 5
  )

  p <- ggviolin(pheno450k_allCE, x = "CellType", y = "RPC", fill = "PTSDlife", color = "PTSDlife", palette = ptsdpal, facet.by = "Sex", alpha = 0.5, xlab = "Leukocyte Subtype", legend.title = "PTSD Status", position = "dodge")  +
    stat_compare_means(aes(group = PTSDlife, label = ..p.signif.., size = 36), method = "wilcox.test", paired = FALSE,  hide.ns = TRUE, label.y.npc = "top") +
    xlab("") +
    ylim(min(pheno450k_allCE$RPC), max(pheno450k_allCE$RPC)) +
    theme_pubr(border = TRUE, legend = "right", x.text.angle = 50) +
    theme(
      strip.text = element_text(face = "bold", size = rel(1.2)),
      axis.title.y = element_text(face = "bold", margin = margin(0, 0.5, 0, 0, unit = "cm")), 
        # legend.position = "right",
      legend.title = element_text(face = "bold"))  
  
print(p)
  
dev.off()

```


```{r fig3-fig5-density-mono-bySex-fig-ggpubr}

## extra page for some reason
CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate(PTSDlife = factor(PTSDlife, labels = c("No PTSD", "PTSD")))

rpc_ptsdlifed <- ggdensity(CombPheno2wcue2, x = "RPC_Mono", color = "PTSDlife", fill = "PTSDlife", palette = ptsdpal, facet.by = "Sex", alpha = 0.4, rug = TRUE, legend.title = "Lifetime PTSD", xlab = FALSE) + theme(strip.text = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 11, face = "plain"),
        legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.position = "top")

cp_ptsdlifed <- ggdensity(CombPheno2wcue2, x = "CP_Mono", color = "PTSDlife", fill = "PTSDlife", palette = ptsdpal, facet.by = "Sex", alpha = 0.4, rug = TRUE, legend.title = "Lifetime PTSD", xlab = FALSE) + theme(strip.text = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 11, face = "plain"),
        legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"), legend.position = "top")

CombPheno2wcue3 <- CombPheno2wcue2 %>% mutate(PTSDStatus3 = case_when(!PTSDStatus == PTSDStatus2 ~ NA_character_, TRUE ~ as.character(PTSDStatus))) %>% mutate(PTSDStatus3 = factor(PTSDStatus3, levels = c("No PTSD", "Remitted", "Current"))) %>% filter(!is.na(PTSDStatus3))


rpc_ptsdstatus3d <- ggdensity(CombPheno2wcue3, x = "RPC_Mono", color = "PTSDStatus3", fill = "PTSDStatus3", palette = ptsdpal2, facet.by = "Sex", alpha = 0.4, rug = TRUE, legend.title = "PTSD Status", xlab = "RPC Mono") + theme(strip.text = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 11, face = "plain"),
        legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.position = "top", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))

cp_ptsdstatus3d <- ggdensity(CombPheno2wcue3, x = "CP_Mono", color = "PTSDStatus3", fill = "PTSDStatus3", palette = ptsdpal2, facet.by = "Sex", alpha = 0.4, rug = TRUE, legend.title = "PTSD Status", xlab = "CP Mono")  + theme(strip.text = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 11, face = "plain"),
        legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.position = "top", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))




pdf(file = paste0(out_dir,"Fig3Raw_Mono_DensityPlot_PTSDlife.pdf"), bg = "transparent", width = 6.5, height = 7.5)

print(ggarrange(rpc_ptsdlifed, cp_ptsdlifed, nrow = 2, legend = "top"))
  
dev.off()


pdf(file = paste0(out_dir,"Fig5Raw_Mono_DensityPlot_PTSDStatus.pdf"), bg = "transparent", width = 6.5, height = 7.5)

print(ggarrange(rpc_ptsdstatus3d, cp_ptsdstatus3d, nrow = 2, legend = "top"))
  
dev.off()

  
```


Figure 4 raw files were derived separately in emmeans analysis folder.  

## Save

```{r date-git}

cat("\n \n This iteration was run on ", processing_date, ".  SHA1 hash for last git commit is ", git, ". \n \n ", sep = "")

```

```{r sessionInfo}

save.image(paste0(out_dir, proj_label, ".RData"))

sessionInfo()

```

