---
title: "Figure Revisions - Combined DNHS+GTP 450K Cell Estimates"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---


```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

# rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(GGally)
library(data.table)
library(ggpubr)
library(ggsci)
library(grid)
library(gtable)
library(gridExtra)
library(RColorBrewer)
library(scales)

```

Generates some additional plots and figures, not in original set generated in **Combined_LeukoComp_450K_FigPlots.Rmd**.  
Load saved image from original set.  

```{r loadData-old-image}

# load("E://GitHub Repositories/gsk-1-SexDiff-Leukocyte/Combined_LeukoComp/20181119_Combined_LeukoComp_450K_Figures_Plots_930ea9a/Combined_LeukoComp_450K_Figures_Plots.RData")

HM450KDat_path <- "D:/Data/DNHS_Data/DNAm_450K/"
load(file = paste0(HM450KDat_path, "CombPheno2wcue2_incTuk.Rdata"))

source("../ModelFxns.R")
source("../PlotFxns.R")

table(CombPheno2wcue2$PTSDlife, CombPheno2wcue2$Sex)

```

```{r out_dir}

proj_label <- "FigRevisions"
processing_date <- format(Sys.time(), "%Y%m%d")
#' git SHA1 hash of last commit.
git <- system("git rev-parse --short HEAD", intern = TRUE)
#' output directory for process and plots
out_dir <- paste0(processing_date,"_",proj_label,"_",git,"/")
print(out_dir)

if (!file.exists(out_dir)) {
dir.create(out_dir,showWarnings = FALSE,recursive = FALSE)
}

```

> Revisions for Figure 1, Supp Fig 1 and 2, to address Adriana's comment. 
> Revisions for Fig 5 and Supp Fig 5 to address using coding of PTSD status, based on CAPS only.  


```{r CombPheno2wcue2-rmpheno}

cells <- CombPheno2wcue2 %>% dplyr::select(c(ends_with("_Mono"), ends_with("_Gran"), ends_with("_CD4T"), ends_with("_CD8T"), ends_with("_NK"), ends_with("_B", ignore.case = FALSE))) %>% dplyr::select(-starts_with("CBS_")) %>% colnames()

CombPheno2wcue2 <- CombPheno2wcue2 %>% mutate_at(vars(cells), funs(.*100)) %>% filter(!is.na(PTSDlife)) 
# CombPheno2wcue2 <- CombPheno2wcue2 %>% filter(!is.na(PTSDlife)) %>% filter(!Basename %in% knownhiv$Basename)
CombPheno2wcue2

```

- final set. v3. 
- scale everything as percent.

```{r set-palette}

sexpalbrewed <- brewer.pal(8, "Dark2")
sexpal <- sexpalbrewed[c(2,3)]

ptsdpal <- c("#619CFF", "#DC0000E5") ## 
ptsdpal2 <- c("#619CFF", "#00BA38", "#E64B35E5") ## 

```


```{r pheno450k_allCE}

rpc <- CombPheno2wcue2 %>% dplyr::select(c(starts_with("RPC_"), matches("SampleID")))
colnames(rpc) <- gsub(colnames(rpc), pattern = "RPC_", replacement = "")

vars <- CombPheno2wcue2 %>% dplyr::select(-c(starts_with("RPC_"), starts_with("CBS_")))
colnames(vars) <- gsub(colnames(vars), pattern = "CP_", replacement = "")

vars.dtl <- melt(data.table(vars), 
                 # id.vars = "SampleID", 
                 measure.vars = c("B", "NK", "CD4T", "CD8T", "Gran", "Mono"), 
                 variable.name = "CellType", 
                 value.name = "CP")

rpc.dtl <- melt(data.table(rpc), 
                 id.vars = "SampleID", 
                 measure.vars = c("B", "NK", "CD4T", "CD8T", "Gran", "Mono"), 
                 variable.name = "CellType", 
                 value.name = "RPC")

setkey(rpc.dtl, SampleID, CellType)
setkey(vars.dtl, SampleID, CellType)

pheno450k_allCE <- merge(vars.dtl,rpc.dtl,by = c("SampleID","CellType"))
pheno450k_allCE

```


```{r suppfig1-pheno450k_allCE-plotcor-spearman}

pdf(file = paste0(out_dir, "SuppFig1Raw_CorComp_bySex_spearman.pdf"), width = 9, height = 5, title = paste0("SuppFig1_CorComp_spearman"), bg = "transparent")

ggscatter(data = pheno450k_allCE, x = "CP", y = "RPC", color = "Sex", palette = sexpal, point = TRUE, facet.by = "CellType", panel.labs = list(CellType = c("CD19+ B cells", "CD56+ NK cells", "CD4+ T cells", "CD8+ T cells", "Granulocytes", "CD14+ monocytes")), add = "reg.line", conf.int = TRUE, size = 1) +
  stat_cor(aes(color = Sex), method = "spearman") +
  geom_abline(slope = 1, intercept = 0) +
  theme(strip.text = element_text(size = 14, face = "bold"),    text = element_text(size = 14, face = "bold"),
legend.title = element_text(face = "bold"), axis.title = element_text(face = "bold"), axis.text = element_text(face = "plain", size = 12), legend.position = "right", axis.title.x = element_text(margin = margin(0.5, 0, 0, 0, unit = "cm")), axis.title.y = element_text(margin = margin(0, 0.5, 0, 0, unit = "cm")))
# , plot.tag = element_text(face = "bold")
dev.off()


## tried making subscript. doesn't work and didn't bother wasting any more time.

# labels1 <- c('CD19^+{}~B cells', 'CD56^+{}~NK~cells', 'CD4^+{}~T cells', 'CD8^+{}~T cells', 'Granulocytes', 'CD14^+{}~monocytes')
# 
# variableLabs <- c(expression(paste(CD19^+{},"B cells")), expression(paste(CD56^+{},"NK cells")), expression(paste(CD4^+{},"Tcells")), expression(paste(CD8^+{},"T cells")), "Granulocytes", expression(paste(CD14^+{},"monocytes")))
# variableLabs <- c(expression(CD19^+{} ~ 'B cells'), expression(CD19^+{} ~ 'B cells'), expression(CD19^+{} ~ 'B cells'), expression(CD19^+{} ~ 'B cells'), "Granulocytes", expression(CD19^+{} ~ 'B cells'))

## works here. plot(CombPheno2wcue2$PC2, CombPheno2wcue2$PC3, xlab = expression(CD19^+{} ~ 'B cells'))

# ggscatter(data = pheno450k_allCE, x = "CP", y = "RPC", color = "Sex", palette = sexpal, point = TRUE, facet.by = "CellType", panel.labs = list(CellType = label_parsed(labels = labels1)), add = "reg.line", conf.int = TRUE, size = 1) +
  # stat_cor(aes(color = Sex), method = "spearman") +
  # geom_abline(slope = 1, intercept = 0) +
  # theme(strip.text = element_text(size = 14, face = "bold"),    text = element_text(size = 14, face = "bold"),
# legend.title = element_text(face = "bold"), axis.title = element_text(face = "bold"), axis.text = element_text(face = "plain", size = 12), legend.position = "right", axis.title.x = element_text(margin = margin(0.5, 0, 0, 0, unit = "cm")), axis.title.y = element_text(margin = margin(0, 0.5, 0, 0, unit = "cm"))) 

# + facet_grid(. ~ CellType, rows = 2, cols = 3)
# panel.labs
# + facet_grid(. ~ CellType, rows = 2, cols = 3)

# , labeller = label_parsed(labels = labels1)
# + label_parsed(variableLabs)
# , facet.by = "CellType", panel.labs = list(CellType = variableLabs)

 # expression(CD19^"+"), expression(CD19^"+"), expression(CD19^"+"),"Granulocytes", expression(CD14^"+")))

# "CD19\u002B B cells", "CD56^{+} NK cells", "CD4^+{} T cells", "CD8^{+} T cells", "Granulocytes", "CD14^{+} monocytes"
```

Spearman makes the most sense. 
Spearman rank correlation: Spearman rank correlation is a non-parametric test that is used to measure the degree of association between two variables.  The Spearman rank correlation test does not carry any assumptions about the distribution of the data and is the appropriate correlation analysis when the variables are measured on a scale that is at least ordinal.

```{r fig1-suppfig2-densityPlots-fig1-ggpubr}

pdf(file = paste0(out_dir, "Fig1Raw_RPC_DensityPlot_byCellType.pdf"), width = 9, height = 5, title = paste0("Fig1Raw_RPC_Density_byCellType"), bg = "transparent")

p1 <- ggdensity(pheno450k_allCE, x = "RPC", fill = "Sex", color = "Sex", palette = sexpal, facet.by = "CellType", panel.labs = list(CellType = c("CD19+ B cells", "CD56+ NK cells", "CD4+ T cells", "CD8+ T cells", "Granulocytes", "CD14+ monocytes")), alpha = 0.5) + theme(strip.text = element_text(face = "bold", size = 12), legend.text = element_text(size = 11, face = "plain"), legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"), axis.text = element_text(size = 12, face = "plain"), legend.position = "bottom", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))

print(p1)

dev.off()

pdf(file = paste0(out_dir,"SuppFig2Raw_CP_DensityPlot_byCellType.pdf"), width = 9, height = 5, title = paste0("SuppFig2Raw_CP_DensityPlot_byCellType"), bg = "transparent")

p2d <- ggdensity(pheno450k_allCE, x = "CP", fill = "Sex", color = "Sex", palette = sexpal, facet.by = "CellType", panel.labs = list(CellType = c("CD19+ B cells", "CD56+ NK cells", "CD4+ T cells", "CD8+ T cells", "Granulocytes", "CD14+ monocytes")), alpha = 0.5) + theme(strip.text = element_text(face = "bold", size = 12), legend.text = element_text(size = 11, face = "plain"), legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"), axis.text = element_text(size = 12, face = "plain"), legend.position = "bottom", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))

print(p2d)

dev.off()

# removed options. add = mean, rug = TRUE

```


```{r fig3-fig5-density-mono-bySex-fig-ggpubr}

CombPheno2wcue3 <- CombPheno2wcue2 %>% filter(!is.na(PTSDStatus))

rpc_PTSDStatusd <- ggdensity(CombPheno2wcue3, x = "RPC_Mono", color = "PTSDStatus", fill = "PTSDStatus", palette = ptsdpal2, facet.by = "Sex", alpha = 0.4, rug = TRUE, legend.title = "PTSD Status", xlab = "RPC Mono") + theme(strip.text = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 11, face = "plain"),
        legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.position = "top", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))

cp_PTSDStatusd <- ggdensity(CombPheno2wcue3, x = "CP_Mono", color = "PTSDStatus", fill = "PTSDStatus", palette = ptsdpal2, facet.by = "Sex", alpha = 0.4, rug = TRUE, legend.title = "PTSD Status", xlab = "CP Mono")  + theme(strip.text = element_text(face = "bold", size = 12),
        legend.text = element_text(size = 11, face = "plain"),
        legend.title = element_text(face = "bold", size = 12), axis.title = element_text(size = 12, face = "bold"),
        axis.text = element_text(size = 12, face = "plain"),
        legend.position = "top", axis.title.x = element_text(margin = margin(0.23, 0, 0, 0, unit = "cm")))




pdf(file = paste0(out_dir,"Fig5Raw_Mono_DensityPlot_PTSDStatus.pdf"), bg = "transparent", width = 6.5, height = 7.5)

print(ggarrange(rpc_PTSDStatusd, cp_PTSDStatusd, nrow = 2, legend = "top"))
  
dev.off()

CombPheno2wcue3
  
```

There are 483 participants in CAPS based coding of PTSD status.  

## Save

```{r date-git}

cat("\n \n This iteration was run on ", processing_date, ".  SHA1 hash for last git commit is ", git, ". \n \n ", sep = "")

```

```{r sessionInfo}

# save.image(paste0(out_dir, proj_label, ".RData"))

sessionInfo()

```

