---
title: "Leukocyte Composition Analysis of Combined DNHS+GTP 450K Cell Estimates - sex*PTSD 2-way ANCOVA with sqrt-transformed monocyte estimates - follow-up"
author: "Grace Kim"
date: '`r Sys.Date()`'
output:
  html_notebook: default
  html_document: default
  pdf_document: 
    includes:
        in_header:table-template.tex
---

```{r knitrOpt, message = FALSE}

library(knitr) # general-purpose package for dynamic report generation in R.
library(rmarkdown)
library(xtable) # export tables to LaTeX or html
library(shiny)
library(stargazer)
library(tools)

rm(list = ls())

opts_knit$set(concordance = TRUE, self.contained = FALSE)

opts_chunk$set(
  error = FALSE,
  tidy = TRUE,
  warning = FALSE,
  highlight = TRUE,
  cache = TRUE,
  comment = NA,
  autodep = TRUE,
  dev = c('png', 'pdf'),
  fig.align = 'center',
  fig.path = 'figure/',
  fig.show = 'asis',
  dpi = 300,
  echo = TRUE,
  out.extra = "keepaspectratio"
) # turn on cache for increased performance and improved styling.

options(stringsAsFactors = FALSE, width = 70, digits = 4)
render_listings()

library(tidyverse)
library(GGally)
# library(data.table)
library(psych)
library(rcompanion)
library(car)          # regression functions (continuous)
library(sandwich) ## heteroskedasticity robust standard errors.  
library(lmtest)   ## lm test
library(nortest) ## normality test
## further list of useful libraries in RegDiagNotes.Rmd

library(MASS) ## negative binomial support
library(lme4)         # random effects
library(goft)
library(jtools)
library(Gmisc)
library(multcompView)
library(Rmisc)
library(ggpubr)
library(ggfortify)

library(sjstats)
library(emmeans)

```

Follow-up on 2 way ANCOVA

```{r load, message = FALSE}

load("Combined_LeukoCompLM_450K_sqrt_Mono100.Rdata")

```


- ANCOVA with 3 digits intead of 5.
adjust manually as necessary for small numbers

```{r ancova-rpc}

options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(anova_stats(RPCancova, digits = 3)[-1,c(1,2,4,3,5,6,8)], format = "pandoc", caption = "ANCOVA table", row.names = FALSE)

```

```{r ancova-cp}
 
options(knitr.kable.NA = '') # this will hide missing values in the kable table
kable(anova_stats(CPancova, digits = 3)[-1,c(1,2,4,3,5,6,8)], format = "pandoc", caption = "ANCOVA table", row.names = FALSE)

```

n=469, b/c this group also was filtered for thosew ithout smoking data. 

```{r describe-rpcmono}

rpc_desc <- dplyr::group_by(data2, Sex, PTSDlife) %>%
  summarise(
    count = n(),
    mean = mean(RPC_Mono, na.rm = TRUE),
    # sd = sd(RPC_Mono, na.rm = TRUE), 
    se = sd(RPC_Mono, na.rm = TRUE)/sqrt(n())
  ) 
# rpc_desc
# 

# merge descriptives with emmeans

rpc_emm <- data.frame(RPC_cld_orig) %>% dplyr::select(-c(one_of(".group", "df")))
# rpc_emm  

rpc_desc2 <- left_join(rpc_desc, rpc_emm, by = c("Sex", "PTSDlife")) %>% rename(EMM = response)
rpc_desc2 %>% kable(., format = "pandoc", caption = "RPC Monocyte by Group")

```

```{r describe-cpmono}

CP_desc <- dplyr::group_by(data2, Sex, PTSDlife) %>%
  summarise(
    count = n(),
    mean = mean(CP_Mono, na.rm = TRUE),
    # sd = sd(CP_Mono, na.rm = TRUE), 
    se = sd(CP_Mono, na.rm = TRUE)/sqrt(n())
  ) 
# CP_desc
# 

# merge descriptives with emmeans

CP_emm <- data.frame(CP_cld_orig) %>% dplyr::select(-c(one_of(".group", "df")))
# CP_emm  

CP_desc2 <- left_join(CP_desc, CP_emm, by = c("Sex", "PTSDlife")) %>% rename(EMM = response)
CP_desc2 %>% kable(., format = "pandoc", caption = "CP Monocyte by Group")

```

Summary for monocyte estimates

Notes on descriptive statistic tables
- EMM stands for estimated marginal means (i.e., least squares means).   
- SE (in capital letters) refers to standard errors for EMM, while se (in lowercase) refers to standard errors for regular means.  
- Lower and upper confidence limits are for 95% level.  
- EMM and intervals were back-transformed from the square-root scale.  
- significance level of alpha = 0.05 was used for EMM comparisons.    
- results for pairwise comparison were averaged over levels for current smoking.  
- degree of freedom was 461 and male lifetime PTSD cases were significantly different from other groups.  


#### RPC emmeans explore

```{r emmeans-rpc-explore}

RPC.mono.emmint

pairs(RPC.emm.s, simple = "each", reverse = TRUE, adjusted = "tukey") ## includes both comparisons. 

# test(contr_rpc, joint = TRUE)

RPC_cld_orig

```

```{r rpc-int2-pairwise}

rpc.int2 <- emmeans(RPCfit, pairwise ~ Sex*PTSDlife)
rpc.int2
contr_rpc2 <- contrast(rpc.int2[[1]], "poly")
contr_rpc2

IC_st <- contrast(rpc.int2[[1]], interaction = c("poly", "consec"))
IC_st

coef(IC_st)

test(IC_st, joint = TRUE)


contrast(emmeans(RPCfit, ~ Sex*PTSDlife), interaction = c("poly", "consec"))

joint_tests(RPCfit)

joint_tests(RPCfit, by = "Sex")

test(contr_rpc2, joint = TRUE)

```

```{r rpc-cld-orig2-emmeans}

RPC_cld_orig2 <- CLD(rpc.int2$emmeans, alpha = 0.05, type = "response", Letters = letters)
###  Remove spaces in .group  
RPC_cld_orig2$.group = gsub(" ", "", RPC_cld_orig2$.group)
RPC_cld_orig2

```

pval adj : tukey method for comparing a family of 4 estimates. 
```{r rpc-cld-orig2-contrasts}

CLD(rpc.int2$contrasts, alpha = 0.05, type = "response", Letters = letters)
###  Remove spaces in .group  
RPC_cld_orig2$.group = gsub(" ", "", RPC_cld_orig2$.group)
RPC_cld_orig2

```


#### CP emmeans explore

```{r emmeans-CP-explore}

CP.mono.emmint

pairs(CP.emm.s, simple = "each", reverse = TRUE, adjusted = "tukey") ## includes both comparisons. 

# test(contr_CP, joint = TRUE)

CP_cld_orig

```

```{r CP-int2-pairwise}

CP.int2 <- emmeans(CPfit, pairwise ~ Sex*PTSDlife)
CP.int2
contr_CP2 <- contrast(CP.int2[[1]], "poly")
contr_CP2

IC_st <- contrast(CP.int2[[1]], interaction = c("poly", "consec"))
IC_st

coef(IC_st)

test(IC_st, joint = TRUE)


contrast(emmeans(CPfit, ~ Sex*PTSDlife), interaction = c("poly", "consec"))

joint_tests(CPfit)

joint_tests(CPfit, by = "Sex")

test(contr_CP2, joint = TRUE)

```

```{r CP-cld-orig2-emmeans}

CP_cld_orig2 <- CLD(CP.int2$emmeans, alpha = 0.05, type = "response", Letters = letters)
###  Remove spaces in .group  
CP_cld_orig2$.group = gsub(" ", "", CP_cld_orig2$.group)
CP_cld_orig2

```

pval adj : tukey method for comparing a family of 4 estimates. 
```{r CP-cld-orig2-contrasts}

CLD(CP.int2$contrasts, alpha = 0.05, type = "response", Letters = letters)
###  Remove spaces in .group  
CP_cld_orig2$.group = gsub(" ", "", CP_cld_orig2$.group)
CP_cld_orig2

```



## Decided not to use. 

```{r hmisc-cindex, message = FALSE, eval = FALSE, include = FALSE}

## decided not to use. 

library(Hmisc)
attach(data2)

male <- data2 %>% filter(Sex == "Male")
fem <- data2 %>% filter(Sex == "Female")

attach(male)
somers2(RPC_Mono, PTSDlife == "no")
somers2(RPC_Mono, PTSDlife == "yes")

attach(fem)
somers2(RPC_Mono, PTSDlife == "no")
somers2(RPC_Mono, PTSDlife == "yes")


# male <- data2 %>% filter(Sex == "Male") %>% mutate(PTSD = case_when(PTSDlife == "no" ~ 0, PTSDlife == "yes" ~ 1))
# fem <- data2 %>% filter(Sex == "Female") %>% mutate(PTSD = case_when(PTSDlife == "no" ~ 0, PTSDlife == "yes" ~ 1))
# 
# somers2(male$RPC_Mono, male$PTSD)
# somers2(fem$RPC_Mono, fem$PTSD)


## notes. 
# consider presenting c-index? concordance probability like auc
# c-index is tightly related to mann whitney. 
# computed from mean of the ranks. 
# pretty similar to accuracy.. 
# somers cor would need some explanation. 

```

- more for effect size.
- decided not to use cohen's d.  similar to delta EMM/beta estimates in EMM comparisons.

```{r cohend, eval = FALSE, include = FALSE}

CombPheno2wcue3_f <- CombPheno2wcue3 %>% filter(Sex == "Female")
CombPheno2wcue3_m <- CombPheno2wcue3 %>% filter(Sex == "Male")

## on untransformed. 

cohen.d(CombPheno2wcue3_f[,c("PTSDlife", "RPC_Mono", "CP_Mono")], group = "PTSDlife")

cohen.d(CombPheno2wcue3_m[,c("PTSDlife", "RPC_Mono", "CP_Mono")], group = "PTSDlife")

```

